## Discrete Probability

We will now transition to probability and statistical inference. We start by covering some basic principles related to categorical data. The subset of probability is referred to as _discrete probability_. It will help us understand the probability theory we will later introduce for numeric and continuous data, which is much more common in data science applications. Discrete probability is more useful in card games and other games of chance and we use these as examples.

### Relative Frequency

The word probability is used in everyday language. For example, Google's auto-complete of "What are the chances of" give us "getting pregnant", "having twins", and "rain today". Answering questions about probability is often hard if not impossible. Here we discuss a mathematical definition of _probability_ that does permit us to give precise answers to certain questions. 

For example, if I have 2 red beads and 3 blue beads inside an urn and I pick one at random, what is the probability of picking a red one? Our intuition tells us that the answer is 2/5 or 40%. A precise definition can be given by noting that there are five possible outcomes of which two satisfy the condition necessary for the event "pick a red bead".  Because each of the five outcomes has the same chance of occurring we conclude that the probability is 0.4 for red and 0.6 for blue. 

A more tangible way to think about the probability of an event is as the proportion of times the event occurs when we repeat the experiment over and over, independently, and under the same conditions. 

### Notation

We use the notation $\mbox{Pr}(A)$ to denote the probability of event $A$ happening. We use the very general term _event_ to refer to things that can happen when something happens by chance. For example, in our previous example the event was "picking a red bead". In a political poll in which we call 100 likely voters at random, an example of an event is "calling 48 Democrats and 52 Republicans".

In data science applications, we will often deal with continuous variables. In these cases events will often be things like "is this person taller than 6 feet". In this case we write events in a more mathematical form: $X \geq 6$. We will see more of these examples later. Here we focus on categorical data.


### Monte Carlo Simulations

Computers provide a way to actually perform the simple random experiment described above: pick a bead at random from a bag with three blue beads and two red ones. Random number generators permit us to mimic the process of picking at random. 

An example is the `sample` function in R. We demonstrate its use in the code below.  First, we use the function `rep` to generate the urn:

```{r}
beads <- rep(c("red", "blue"), times = c(2,3))
beads
```

and then use `sample` to pick a bead at random:

```{r}
sample(beads, 1)
```

This line of code produces one random outcome. We want to repeat this experiment "over and over". However, it is of course impossible to repeat forever. Instead, we repeat the experiment a large enough number of times to make the results practically equivalent.  This is an example of a _Monte Carlo_ simulation. 

Note that much of what mathematical and theoretical statisticians study, something we do not cover in this course, relates to providing rigorous definitions of "practically equivalent" as well as studying how close a large number of experiments gets us to what happens in the limit. Later in this lecture we provide a practical approach to deciding what is "large enough".

To perform our first Monte Carlo simulation we use the `replicate` function, which permits us to repeat the same task any number of times. Here we repeat the random event $B=$ 10,000 times:

```{r}
B <- 10000
events <- replicate(B, sample(beads, 1))
```

We can now see if in fact, our definition is in agreement with this Monte Carlo simulation approximation. We can use `table` to see the distribution:

```{r}
tab <- table(events)
tab
```

and `prop.table` gives us the proportions:

```{r}
prop.table(tab)
```

The numbers above are the estimated probabilities provided by this Monte Carlo simulation. Statistical theory, not covered here, tells us the as $B$ gets larger, the estimates get closer to 3/5 = 0.6 for blue and 2/5 = 0.4 for red. 

This is a simple and not very useful example, but we will use Monte Carlo simulation to estimate probabilities in cases in which it is harder to compute the exact ones. Before we go into more complex examples we use simple ones to demonstrate the computing tools available in R.

### With and without replacement

The function `sample` has an argument that permits us to pick more than one element from the urn. However, by default, this selection occurs _without replacement_: after a bead is selected, it is not put back in the bag. Note what happens when we ask to randomly select five beads:

```{r}
sample(beads, 5)
sample(beads, 5)
sample(beads, 5)
```

This results in re-arrangements that always have three blue and two red beads. If we ask that six beads be selected, we get an error

```{r, eval=FALSE}
sample(beads, 6)
```

`Error in sample.int(length(x), size, replace, prob) : 
  cannot take a sample larger than the population when 'replace = FALSE'`

However, the `sample` function can be used directly, without the use of `replicate`, to repeat the same experiment of picking one out of the 5 beads, over and over, under the same conditions. To do this we sample _with replacement_: return the bead back to the urn after selecting it. 

We can tell `sample` to do this by changing the `replace` argument, which defaults as `FALSE`, to  `replace = TRUE`:

```{r}
events <- sample(beads, B, replace = TRUE)
prop.table(table(events))
```

Note that, not surprisingly, we get results very similar to 
those previously obtained with `replicate`.
  
### Probability Distributions

Defining a distribution for categorical outcomes is relatively straight forward. We simply assign a probability to each category.
In cases that can be thought of as beads in an urn, for each bead type their proportion defines the distribution. 

If we are randomly calling likely voters from a population that is 44% Democrat, 44% republican, 10% undecided and 2% green party, these proportions define the  probability for each group. The probability distribution is:

$$
\mbox{Pr}(\mbox{picking a Republican})=0.44\\ \mbox{Pr}(\mbox{picking a Democrat})=0.44\\
\mbox{Pr}(\mbox{picking an undecided})=0.10\\
\mbox{Pr}(\mbox{picking a Green})=0.02\\
$$

### Independence

We say two events are independent if the outcome of one does not affect the other. The classic example are coin tosses. Every time we toss a fair coin the probability of seeing heads is 1/2 regardless of what previous tosses have revealed. The same is true when we pick beads from an urn with replacement. In the example above the probability of red is 0.40 regardless of previous draws. 

Many examples of events that are not independent come from card games. When we deal the first card, the probability of getting a king is 1/13 since there are 13 possibilities: Ace, Deuce, Three, $\dots$, Ten, Jack, Queen, and King, and there are 4 of each possibility (for the 4 suits hearts, spades, diamonds, and clubs), totaling 52 cards. Now if we deal a king for the first card, and don't replace it into the deck, the probability of a second card being a king is less because there are only three kings left: the probability is 3 out of 51. These events are therefore **not independent**. The first outcome affects the next.

To see an extreme case of non-independent events, consider our example of drawing five beads at random **without** replacement:

```{r, echo=FALSE}
set.seed(1)
```

```{r}
x <- sample(beads, 5)
```

If you have to guess the color of the first bead you predict blue since blue has a 60% chance. But if I show you the result of the last four outcomes:

```{r}
x[2:5]
```

would you still guess blue? Of course not. Now you know that the probability of red is 1 since only a red bead remains. The events are not independent so the probabilities change.

### Conditional Probabilities

When events are not independent, _conditional probabilities_ are useful. We already saw an example of a conditional probability: we computed the probability that a second dealt card is a king given that the first was a king. In probability we use the following notation:

$$
\mbox{Pr}(\mbox{Card 2 is a king} \mid \mbox{Card 1 is a king}) = 3/51
$$

We use the $\mid$ as shorthand for "given that" or "conditional on". 

Note that when two events, say $A$ and $B$, are independent we have

$$
\mbox{Pr}(A \mid B) = \mbox{Pr}(A) 
$$

This is the mathematical way of saying: the fact that $B$ happened does not affect the probability of $A$ happening.
In fact, this can be considered the mathematical definition of independence.

## Multiplication rule

If we want to know the probability of two events, say $A$ and $B$, occurring, we can use the multiplication rule.

$$ 
\mbox{Pr}(A \mbox{ and } B) = \mbox{Pr}(A)\mbox{Pr}(B \mid A)
$$
Let's use Black Jack as an example. In Black Jack you get assigned two random cards. After you see what you have, you can ask for more. The goal is to get closer to 21 than the dealer, without going over. Number cards 2-10 are worth their number in points, face cards (Jacks, Queens, Kings) are worth 10 points and aces worth 11 or 1 (you choose).

So, in a black jack game, to calculate the chances of getting a 21 in the following way by drawing an ace and then a face card, we compute the probability of the first being an ace and multiply by the probability of a face card given that the first was an ace: $1/13 \times 12/51 \approx 0.018$

The multiplicative rule also applies to more than two events. We can use induction to expand for more events:

$$ 
\mbox{Pr}(A \mbox{ and } B \mbox{ and } C) = \mbox{Pr}(A)\mbox{Pr}(B \mid A)\mbox{Pr}(C \mid A \mbox{ and } B)
$$

#### Multiplication rule under independence 

When we have independent events then the multiplication rule becomes simpler:

$$ 
\mbox{Pr}(A \mbox{ and } B \mbox{ and } C) = \mbox{Pr}(A)\mbox{Pr}(B)\mbox{Pr}(C)
$$

But we have to be very careful before using this, as assuming independence can result in very different, and incorrect, probability calculations when we don't actually have independence.

As an example, imagine a court case in which the suspect was described to have a mustache and a beard. The defendant has a mustache and a beard and the prosecution brings in an "expert"  to testify that 1/10 men have beards and 1/5 have mustaches so using the multiplication rule we conclude that only $1/10 \times 1/5$ or 0.02 have both. 

But to multiply like this we need to assume independence! The conditional probability of a man having a mustache conditional on them having a beard is .95. So the correct calculation probability is much higher: 0.09.


Note that the multiplication rule also gives us a general formula for computing conditional probabilities:


$$ 
\mbox{Pr}(B \mid A) = \frac{\mbox{Pr}(A \mbox{ and } B)}{ \mbox{Pr}(A)}
$$

To illustrate how we use these formulas and concepts in practice we will show several examples related to card games.

### Combinations and Permutations

In our very first example we imagined an urn with five beads. Let's review how we did this. To compute the probability distribution of one draw, we simply listed out all the possibilities, there were 5, and then, for each event, counted how many of these possibilities were associated with the event. So, for example, because out of the five possible outcomes, three were blue, the probability of blue is 3/5. 

For more complicated examples these computations are not straightforward. For example, what is the probability that if I draw five cards without replacement I get all cards of the same suit; what is called a flush in poker? In a Discrete Probability course you learn theory on how to make these kinds of computations. Here we focus on how to use R code to compute the answers.

First let's construct a deck of cards. For this we will use the `expand.grid`  and `paste` functions. We use `paste` to create strings by joining smaller strings. For example, if we have the number and suit of a card we create the card name like this:

```{r}
number <- "Three"
suit <- "Hearts"
paste(number, suit)
```

`paste` also works on pairs of vectors performing the operation element-wise:

```{r}
paste(letters[1:5], as.character(1:5))
```

The function `expand.grid` gives us all the combinations of entries of two vectors. For example if you have blue and black pants and white, grey and plaid shirts all your combinations are:

```{r}
expand.grid(pants = c("blue", "black"), shirt = c("white", "grey", "plaid"))
```

So here is how we generate a deck of cards:
```{r}
suits <- c("Diamonds", "Clubs", "Hearts", "Spades")
numbers <- c("Ace", "Deuce", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nine", "Ten", "Jack", "Queen", "King")
deck <- expand.grid(number=numbers, suit=suits)
deck <- paste(deck$number, deck$suit)
```

With the deck constructed, we can now double check that the probability of drawing a king as the first card is 1/13. We simply compute the proportion of possible outcomes that satisfy our condition:

```{r}
kings <- paste("King", suits)
mean(deck %in% kings)
```

which is 1/13. 

Now, how about the conditional probability of the second card being a king given that the first was a king ? Earlier we deduced that if one king is already out of the deck and there are 51 left then this probability is 3/51.
Let's confirm by listing out all possible outcomes. 

To do this we can use the `permutations` function from the `gtools` package. This function computes, for any list of size `n`, all the different combinations we can get when we select `r` items. So here are all the ways we can chose 2 numbers from a list consisting of `1,2,3`:

```{r}
library(gtools)
permutations(3, 2)
```

Notice that the order matters here: 3, 1 is different than 1,3. Also note that (1,1), (2,2) and (3,3) don't appear because once we pick a number it can't appear again.

Optionally, we can add a vector. So if you want to see five random seven digit phone numbers, out of all possible phone numbers you could type:

```{r, eval=FALSE}
all_phone_numbers <- permutations(10, 7, v = 0:9)
n <- nrow(all_phone_numbers)
index <- sample(n, 5)
all_phone_numbers[index,]
```

Instead of using the numbers 1 through 10, the default, it uses what we provided through `v`: the digits 0 through 9.

To compute all possible ways we can chose two cards, when the order matters, we type:
```{r}
hands <- permutations(52, 2, v = deck)
```

This is a matrix with two columns and `r nrow(hands)` rows. With a matrix we can get the first and second card like this:

```{r}
first_card <- hands[,1]
second_card <- hands[,2]
```

Now the cases for which the first card was a king can be computed like this:

```{r}
kings <- paste("King", suits)
sum(first_card %in% kings)
```

To get the conditional probability we compute what fraction of these have a king in the second card:

```{r}
sum(first_card %in% kings & second_card %in% kings) / sum(first_card %in% kings)
```

which is exactly 3/51 as we had already deduced. Note that the code above is equivalent to 

```{r}
mean(first_card %in% kings & second_card %in% kings) / mean(first_card %in% kings)
```

which uses `mean` instead of `sum` and is an R version of 

$$
\frac{\mbox{Pr}(A \mbox{ and } B)}{ \mbox{Pr}(A)}
$$


Now what if the order does not matter? For example, in Black Jack if you get an Ace and face card in the first draw it is called a _Natural 21_ and you win automatically. If we want to compute the probability of this happening we want to enumerate the _combinations_, not the permutations, since the order does not matter. Note the differences:

```{r}
permutations(3,2)
combinations(3,2)
```

In the second line the outcome does not include (2,1) because the (1,2) already was enumerated. Similarly for (3,1) and (3,2).

So to compute the probability of a _Natural 21_ in Blackjack we can do this:

```{r}
aces <- paste("Ace", suits)

facecard <- c("King", "Queen", "Jack", "Ten")
facecard <- expand.grid(number = facecard, suit = suits)
facecard <- paste(facecard$number, facecard$suit)

hands <- combinations(52, 2, v = deck)
mean(hands[,1] %in% aces & hands[,2] %in% facecard)
```

Note that in the last line we assume the ace comes first. This is only because we know the way `combination` enumerates possibilities and it will list this case first. But to be safe we could have written this to get the same answer:

```{r}
mean((hands[,1] %in% aces & hands[,2] %in% facecard) | (hands[,2] %in% aces & hands[,1] %in% facecard))
```


#### Monte Carlo Example

Instead of using `combinations` to deduce the exact probability of a Natural 21 we can use a Monte Carlo to estimate this probability. In this case we draw two cards over and over and keep track of how many 21s we get. We can use the function `sample` to draw two cards without replacement:

```{r}
hand <- sample(deck, 2)
hand
```

And then check if one card is an ace and the other a face card or a 10. Going forward we include 10 when we say _face card_. Now we need to check both possibilities:

```{r}
(hands[1] %in% aces & hands[2] %in% facecard) | (hands[2] %in% aces & hands[1] %in% facecard)
```

If we repeat this 10,000 times, we get a very good approximation of the probability of a Natural 21.

Let's start by writing a function that draws a hand and returns TRUE if we get a 21. The function does not need any 
arguments because it uses objects defined in the global environment.
```{r}
blackjack <- function(){
   hand <- sample(deck, 2)
  (hand[1] %in% aces & hand[2] %in% facecard) | 
    (hand[2] %in% aces & hand[1] %in% facecard)
}
```

Note that here we do have to check both possibilities: Ace first or Ace second because we are not using the `combinations` function. The function returns `TRUE` if we get a 21 and `FALSE` otherwise:

```{r}
blackjack()
```

Now we can play this game, say, 10,000 times:

```{r}
B <- 10000
results <- replicate(B, blackjack())
mean(results)
```

### Birthday Problem

Suppose you are in a classroom with 50 people. If we assume this is a randomly selected group of 50 people, what is the chance that at least two people have the same birthday? Although it is somewhat advanced, we can deduce this mathematically. We do this later. Here we use a Monte Carlo simulation. For simplicity, we assume nobody was born on February 29. This actually doesn't change the answer much. 

First note that birthdays can be represented as numbers between 1 and 365, so a sample of 50 birthdays can be obtained like this:

```{r,echo=FALSE}
set.seed(1)
```

```{r}
n <- 50
bdays <- sample(1:365, n, replace = TRUE)
```

To check if in this particular set of 50 people we have at least two with the same birthday we can use the function `duplicated` which returns `TRUE` whenever an element of a vector is a duplicate. Here is an example:

```{r}
duplicated(c(1,2,3,1,4,3,5))
```

The second time 1 and 3 appear we get a `TRUE`. So to check if two birthdays were the same we simply use the `any` and `duplicated` functions like this:

```{r}
any(duplicated(bdays))
```

In this case, we see that it did happen. At least two people had the same birthday.

```{r}
bdays
```


### sapply: a better way to do for loops

Say we want to use this knowledge to bet with friends about two people having the same birthday in a group of people. When are the chances larger than 50%? Larger the 75%? 

Let's create a look-up table.
We can quickly create a function to compute this for any group size:

```{r}
same_birthday <- function(n){
  bdays <- sample(1:365, n, replace=TRUE)
  any(duplicated(bdays))
}

compute_prob <- function(n, B=10000){
  results <- replicate(B, same_birthday(n))
  mean(results)
}
```

And now we can use a for-loop to run it for several group sizes:

```{r}
n <- seq(1,60)
```


Now, for-loops are rarely the preferred approach in R. In general, we try to perform operations on entire vectors. Arithmetic operations, for example, operate on vectors in an element-wise fashion:

```{r}
x <- 1:10
sqrt(x)
y <- 1:10
x*y
```

No need for for-loops. But not all functions work this way. For example, the function we just wrote does not work element-wise since it is expecting a scalar (just one number). This piece of code does not run the function on each entry of `n`:

```{r, eval=FALSE}
compute_prob(n)
```

The function `sapply` permits us to perform element-wise operations on any function. Here is how it works:

```{r}
x <- 1:10
sapply(x, sqrt)
```

It loops through the elements of the first argument of `sapply` and sends those as values to first argument of the function specified as the second argument to `sapply`. So for our case we can simply type:

```{r, eval=FALSE}
prob <- sapply(n, compute_prob)
```

We can now make a plot of the estimated probabilities of two people having the same birthday in a group of size $n$:

```{r}
prob <- sapply(n, compute_prob)
plot(n, prob)
```

Now let's compute the exact probabilities rather than use Monte Carlo approximations. Not only do we get the exact answer using math but the computations are  much faster since we don't have to generate experiments.  


To make the math simpler, instead of computing the probability of it happening we will compute the probability of it not happening. We use the multiplication rule.


### How many Monte Carlo experiments are enough

In the examples above we used $B=$ 10,000 Monte Carlo experiments. It turns out that this provided very accurate estimates. But in more complex calculations, 10,000 may not be nearly enough. Also for some calculations, 10,000 experiments might not be computationally feasible. In practice we won't know what the answer is so we won't know if our Monte Carlo estimate is accurate. We know that the larger $B$ is, the better the approximation. But how big do we need it to be? This is actually a challenging question and answering it often requires advanced theoretical statistics training. 

One practical approach we will describe here is to check for the stability of the estimate. Here is an example with the birthday problem for a group of 25 people. 

```{r}
B <- 10^seq(1, 5, len = 100)
compute_prob <- function(B, n=25){
  same_day <- replicate(B, same_birthday(n))
  mean(same_day)
}
prob <- sapply(B, compute_prob)
plot(log10(B), prob, type = "l")
```

In this plot we can see that the values start to stabilize, (vary less than .01), around 1000. Note that the exact probability, which we know in this case, is:

```{r}
exact_prob <- function(n){
  prob_unique <- seq(365,365-n+1)/365 
  1 - prod( prob_unique)
}
eprob <- sapply(n, exact_prob)
eprob[25]
```


### Addition Rule

Another way to compute the probability of a Natural 21 is to notice that it is the probability of an ace followed by a facecard or a facecard followed by an ace. Here we use the addition rule

$$
\mbox{Pr}(A \mbox{ or } B) = \mbox{Pr}(A) + \mbox{Pr}(B) - \mbox{Pr}(A \mbox{ and } B)
$$

This rule is intuitive: think of a Venn diagram. If we simply add the probabilities we count the intersection twice.
```{r,echo=FALSE, message=FALSE, warning=FALSE}
library(VennDiagram)
grid.newpage()
draw.pairwise.venn(22, 20, 11, category = c("A", "B"), lty = rep("blank", 2), fill = c("light blue", "pink"), alpha = rep(0.5, 2),  cat.dist = rep(0.025, 2), cex=0)
```

In the case of Natural 21 the intersection is empty since both hands can't happen simultaneously. The probability of an ace followed by a face card is $1/13 \times 16/51$ and the probability of a face card followed by an ace is $16/52 \times 4/51$. These two are actually the same which makes sense due to symmetry. In any case we get the same result using the addition rule:

```{r}
1/13*16/51 + 16/52*4/51 - 0
```
### Monty Hall Problem

In the 1970s there was a game show called "Let's Make a Deal". Monty Hall was the host. At some point in the game contestants were asked to pick one of three doors. Behind one door there was a prize. The other doors had a goat to show you had lost. After the contestant chose a door, Monty Hall would open one of the two remaining doors and show the contestant there was no prize.  Then he would ask, "Do you want to switch doors?" What would you do?

We can use probability to show that if you stick to the original door your chances of winning a prize are 1 in 3 but if you switch, your chances double to 2 in 3! This seems counterintuitive. Many people incorrectly think both chances are 1 in 2 since you are choosing between 2. You can watch a detailed explanation [here](https://www.khanacademy.org/math/precalculus/prob-comb/dependent-events-precalc/v/monty-hall-problem) or read one [here](https://en.wikipedia.org/wiki/Monty_Hall_problem). Here we use a Monte Carlo simulation to see which strategy is better. Note that this code is written longer than it should be for pedagogical purposes. 

Let's start with the stick strategy:

```{r}
B <- 10000
stick <- replicate(B, {
  doors <- as.character(1:3)
  prize <- sample(c("car","goat","goat"))
  prize_door <- doors[prize == "car"]
  my_pick  <- sample(doors, 1)
  show <- sample(doors[!doors %in% c(my_pick, prize_door)],1)
  stick <- my_pick
  stick == prize_door
})
mean(stick)
```

As we write the code we note that the lines starting with `my_pick` and `show` have no influence on the last logical operation. From this we should realize that the chance is 1 in 3, what we started out with.

Now let's repeat the exercise but consider the switch strategy:

```{r}
switch <- replicate(B, {
  doors <- as.character(1:3)
  prize <- sample(c("car","goat","goat"))
  prize_door <- doors[prize == "car"]
  my_pick  <- sample(doors, 1)
  show <- sample(doors[!doors %in% c(my_pick, prize_door)], 1)
  stick <- my_pick
  switch <- doors[!doors%in%c(my_pick, show)]
  switch == prize_door
})
mean(switch)
```

The Monte Carlo estimate confirms the 2/3 calculation. This helps us gain some insight by showing that we are removing a door, `show`, that is definitely not a winner from our choices. We also see that unless we get it right when we first pick, you win: 1 - 1/3 = 2/3 of the time. 
