## Combining tables

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
library(dslabs)
ds_theme_set()
```

The information we need for a given analysis may not be just one table. For example, when forecasting elections we used the function `left_join` to combine the information from two tables. Here we use a simpler example to illustrate the general challenge of combining tables.

Suppose we want to explore the relationship between population size for US states, which we have in this table

```{r}
data(murders)
head(murders)
```

and electoral votes, which we have in this one:

```{r}
data(polls_us_election_2016)
head(results_us_election_2016)
```

Notice that just joining these two tables together will not work since the order of the states is not quite the same:

```{r}
identical(results_us_election_2016$state, murders$state)
```

The _join_ functions, described below, are designed to handle this challenge.

### Joins

The `join` functions in the `dplyr` package, which are based on SQL joins, make sure that the tables are combined so that matching rows are together.
The general idea is that one needs to identify one or more columns that will serve to match the two tables. Then a new table with the combined information is returned. Note what happens if we join the two tables above by `state` using `left_join`:

```{r}
tab <- left_join(murders, results_us_election_2016, by = "state")
tab %>% select(state, population, electoral_votes) %>% head()
```

The data has been successfully joined and we can now, for example, make a plot to explore the relationship between population and electoral votes:

```{r}
tab %>% ggplot(aes(population/10^6, electoral_votes, label = abb)) +
  geom_point() +
  geom_text_repel() + 
  scale_x_continuous(trans = "log2") +
  scale_y_continuous(trans = "log2") +
  geom_smooth(method = "lm", se = FALSE) +
	xlab("Population (in millions)") + 
	ylab("Electoral Votes")
```

We see the relationship is close to linear with about 2 electoral votes for every million persons, but with smaller states getting a higher ratio.


In practice, it is not always the case that each row in one table has a matching row in the other. For this reason we have several different ways to join. To illustrate this challenge, take subsets of the matrices above:

```{r}
tab1 <- slice(murders, 1:6) %>% select(state, population)
tab1
```

so that we no longer have the same states in the two tables:
```{r}
tab2 <- slice(results_us_election_2016, c(1:3, 5, 7:8)) %>% select(state, electoral_votes)
tab2
```

We will use these two tables as examples.

#### Left join

Suppose we want a table like `tab1` but adding electoral votes to whatever states we have available. For this we use left join with `tab1` as the first argument.

```{r}
left_join(tab1, tab2)
```

Note that `NA`s are added to the two states not appearing in `tab2`. Also note that this function, as well as all the other joins, can receive the first arguments through the pipe:

```{r}
tab1 %>% left_join(tab2)
```


#### Right join

If instead of a table like `tab1` we want one like `tab2` we can use `right_join`:

```{r}
tab1 %>% right_join(tab2)
```

Notice that now the NAs are in the column coming from `tab1`.

#### inner join

If we want to keep only the rows that have information in both tables we use inner join. You can think of this an intersection:

```{r}
inner_join(tab1, tab2)
```

#### full join

And if we want to keep all the rows, and fill the missing parts with NAs, we can use a full join. You can think of this as a union:

```{r}
full_join(tab1, tab2)
```

#### semi join

The `semi_join` let's us keep the part of the first table for which we have information in the second. It does not add the columns of the second:

```{r}
semi_join(tab1, tab2)
```


#### anti joun

The function `anti_join` is the opposite of `semi_join`. It keeps the elements of the first table for which there is no information in the second:

```{r}
anti_join(tab1, tab2)
```

### Binding

Although we have yet to use it in this course, another common way in which datasets are combined is by _binding_ them. Unlike the join function, the binding functions do no try to match by a variable but rather just combine datasets. If the datasets don't match by the appropriate dimensions one obtains an error.

#### Columns

The `dplyr` function _bind_cols_ binds two objects by making them columns in a tibble. For example, if we quickly want to make a data frame consisting of numbers we can use.

```{r}
bind_cols(a = 1:3, b = 4:6)
```

This function requires that we assign names to the columns. Here we chose `a` and `b`. 

Note there is an R-base function `cbind` that performs the same function but creates objects other than tibbles. 

`bind_cols` can also bind data frames. For example, here we break up the `tab` data frame and then bind them back together:

```{r}
tab1 <- tab[, 1:3]
tab2 <- tab[, 4:6]
tab3 <- tab[, 7:9]
new_tab <- bind_cols(tab1, tab2, tab3)
head(new_tab)
```


#### rows

The `bind_rows` is similar but binds rows instead of columns.

```{r}
tab1 <- tab[1:2,]
tab2 <- tab[3:4,]
bind_rows(tab1, tab2)
```

This is based on an R-base function `rbind`.

### Set Operators

Another set of commands useful for combing are the set operators. When applied to vectors, these behave as their names suggest. However, if the `tidyverse`, or more specifically, `dplyr` is loaded, these functions can be used on data frames as opposed to just on vectors.

#### Intersect

You can take intersections of vectors:

```{r}
intersect(1:10, 6:15)
```

```{r}
intersect(c("a","b","c"), c("b","c","d"))
```

But with `dplyr` loaded we can also do this for tables having the same column names:

```{r}
tab1 <- tab[1:5,]
tab2 <- tab[3:7,]
intersect(tab1, tab2)
```


#### Union

Similarly _union_ takes the union:

```{r}
union(1:10, 6:15)
```

```{r}
union(c("a","b","c"), c("b","c","d"))
```

But with `dplyr` loaded we can also do this for tables having the same column names:

```{r}
tab1 <- tab[1:5,]
tab2 <- tab[3:7,]
union(tab1, tab2)
```


#### Set differrence

The set difference between a first and second argument can be obtained with `setdiff`. Not unlike `instersect` and `union`, this function is not symmetric:


```{r}
setdiff(1:10, 6:15)
setdiff(6:15, 1:10)
```

As with the others above, we can apply it to data frames:
```{r}
tab1 <- tab[1:5,]
tab2 <- tab[3:7,]
setdiff(tab1, tab2)
```

#### `setequal`

Finally, the function `set_equal` tells us if two sets are the same, regardless of order. So

```{r}
setequal(1:5, 1:6)
```

but

```{r}
setequal(1:5, 5:1)
```

When applied to data frames that are not equal regardless of order, it provides a useful message letting us know how the sets are different:

```{r}
setequal(tab1, tab2)
```

