## Data types

Variables in R can be of different types. For example we need to distinguish numbers from character strings and tables from simple lists of numbers. The function `class` helps us determine what type of object we have:

```{r}
a <- 2
class(a)
```

To work efficiently in R it is important to learn the different types of variables and what we can do with these.

### Data Frames

Up to now, the variables we have defined are just one number. This is not very useful for storing data. The most common way of storing a dataset in R is in a _data frame_. Conceptually, we can think of a data frame as table with rows representing observations and the different variables reported for each observatin defining the columns. Data frames are particularly useful for datasets because we can combine different data types into one object. 

We stored the data for our motivating example in a data frame. You can access this dataset by loading the `dslabs` library and loading the `murders` dataset using the `data` function:

```{r}
library(dslabs)
data(murders)
```

To see that this is in fact a data frame we type

```{r}
class(murders)
```

### Examining an object

The function `str` is useful to find out more about the structure of an object

```{r}
str(murders)
```

This tells us much more about the object. We see that the table has 51 rows (50 states plus DC) and five variables. We can show the first six lines using the function `head`:

```{r}
head(murders)
```

In this data set each state is considered an obsrvation and five varialbles reported for each state.

Before we go any further in answering our original question about different states, let's get to know the components of this object better.

### The accessor

For our analysis we will need to access the different variables, represented by columns, included in this data frame. To access these variables  we use the accessor operator `$` in the following way:

```{r}
murders$population
```

But how did we know to use `population`? Above, by applying the function `str` to the obejct `murders`, we revealed the names for each of the five variables stored in this table. We can quickly access the variables names using:

```{r}
names(murders)
```

It is important to know that the order of the entries in `murders$population` preserve the order of the rows in our data table. This will later permit us to manipulate one variable based on the results of another, for example we will be able to order the state names by number of murders.

**Tip**: R comes with a very nice auto-complete functionality that saves us the trouble of typing out all the name. Try typing `murders$p` then hitting the _tab_ key on your keyboard. RStudio has many useful auto-complete features options.

### Vectors: numerics, characters, and logical

Note that the object  `murders$population` is not one number but several. We call these types of objects _vectors_. A single number is technically a vector but in general we vectors refer to objects with several entries. The function `length` tells you how many entries in the vector:

```{r}
pop <- murders$population
length(pop)
```

This particular vector is _numeric_ since population sizes are numbers:

```{r}
class(pop)
```
In a numeric vector, every entry must be a number. 

To store character strings, vectors can also be of class _character_. For example the state names are characters:

```{r}
class(murders$state)
```

As with numeric vectors, all entries in a character verctor need to be a character.

Another important type  are _logical vectors_. These must be either `TRUE` or `FALSE`. 

```{r}
z <- 3 == 2
z
class(z)
```

Here the `==` is a relational operator asking if 3 is equals to 2. Remember that in R, if you just use one `=` you actually assign. You can see the other _relational operators_ by typing

```{r, eval=FALSE}
?Comparison
```

In future sections you will see how useful relational operators can be.


**Advanced**: Mathematically, the values in `pop` are integers and there is an integer class in R. However, by default, numbers are assigned class numeric even when they are round integers, for example `class(1)` returns numeric. You can turn them into class integer with `as.integer(1)` or add by adding an `L` like this: `1L`. Note the class by typing: `class(1L)`


### Factors

In the `murders` dataset we might expect the region to also be a character vector. However it is not:

```{r}
class(murders$region)
```

it is a _factor_. Factors are useful for storing categorical data. Notice that there are only 4 regions:


```{r}
levels(murders$region)
```

So, in the background, R stores these _levels_ as integers and keeps a map to keep track of the labels. This is more memory efficient than storing all the characters. However, factors are also a source of confusion as they can easily be confused with characters but behave differently in different context. We will see more of this later. 

In general, we recommend avoiding factors as much as possible although they are sometimes necessary to fit model containing categorical data.
 
### Lists

Data frames are a special case of _lists_. We will covers lists in more detail later but know that they are useful because you can store any combination of other types, here is an example of list we created for you:


```{r, echo=FALSE}
record <- list(name = "John Doe",
             student_id = 1234,
             grades = c(95, 82, 91, 97, 93),
             final_grade = "A")
```

```{r}
record
class(record)
```

We won't be using lists until later but you might encounter one in your own exploration of R. Note that, as with data frames, you can extract the components with the accessor `$`. In fact, data frames are a type of list.

```{r}
record$student_id
```

We can also use double brackets like this:

```{r}
record[["student_id"]]
```

You should get used to the fact that in R there are several ways to do the same thing, in particular accessing entries.


