---
title: "Assessment Solutions"
output: html_document
---

```{r}
library(dslabs)
data(murders)
```

1. Compute the per 100,000 murder rate for each state and store it in an object called `murder_rate`. Then use the logical operators to create a logical vector, name it `low`, that tells us which entries of `murder_rate` are lower than 1.

```{r}
murder_rate <- murders$total / murders$population * 100000 
low <- murder_rate < 1
low
```

2. Now use the results from the previous exercise and the function `which` to determine the indices of `murder_rate` associated with values lower than 1.

```{r}
ind <- which(murder_rate < 1)
```


3. Use the results from the previous exercise to report the names of the states with murder rates lower than 1.

```{r}
murders$state[ind]
```


4. Now extend the code from exercises 2 and 3 to report the states in the Northeast with murder rates lower than 1. Hint: Use the previously defined logical vector `low` and the logical operator `&`.

```{r}
northeast <- murders$region == "Northeast"
ind <- low & northeast
murders$state[ind]
```

5. In a previous exercise we computed the murder rate for each state and the average of these numbers. How many states are below the average?

```{r}
avg <- mean(murder_rate)
sum(murder_rate < avg)
```


6. Use the match function to identify the states with abbreviations AK, MI, and IA. Hint: Start by defining an index of the entries of `murders$abb` that match the three abbreviations, then use the `[]` operator to extract the states.

```{r}
ind <- match(c("AK", "MI", "IA"), murders$abb)
murders$state[ind]
```

7. Use the `%in%` operator to create a logical vector that answers the question: which of the following are actual abbreviations: MA, ME, MI, MO, MU?

```{r}
c("MA", "ME", "MI", "MO", "MU") %in% murders$abb
```


 8. Extend the code you used in exercise seven to report the one entry that is **not** an actual abbreviation. Hint: Use the `!` operator, which turns `FALSE` into `TRUE` and vice-versa, then `which` to obtain an index.

```{r}
which(c("MA", "ME", "MI", "MO", "MU") %in% murders$abb != TRUE)
```

